---
title: "MA50258 Applied Statistics: Coursework 2"
author: "09618"
output:
  html_document:
    number_sections: no
    toc: yes
    toc_float: yes
    code_folding: hide
    code_download: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

```{r libraries,echo= FALSE}
# include your libraries here
# use echo = FALSE if you do not want the Code button to appear 
library(tidyverse)
library(modelr)
library(readr)
library(here)

# Load external libraries
library(dplyr)
library(janitor)
library(ggplot2)
library(kableExtra)
library(MASS)
library(broom)
library(stats)
```

```{r,message=FALSE}
# read in the data
sdg <- read_csv(here("data", "sdg.csv"))
Wine <- read_csv(here("data", "Wine.csv"))
```



```{=html}
<!--
You can include non-coding comments using these delimiters. What is written inside these delimiters wont be knitted into the html document!

Do not change the structure of the document, just fill in the sections as indicated 
You can only change the headings indicated with [*]
-->
```
# Sustainable development goals (SDG) data

## Cleaning the dataset

```{r}
# Use janitor to clean the data
sdg_clean <- sdg %>%
  clean_names() %>%
  remove_empty(c("rows", "cols")) %>% 
  drop_na()
```

## Performing exploratory data analysis

```{r}
# Check the structure of the data
str(sdg_clean)
```

```{r}
# Summarise the data
summary(sdg_clean)
```


```{r}
# Plot the distribution of infant mortality rate
ggplot(data = sdg_clean, aes(x = infant_mortality_rate)) +
  geom_histogram(aes(y = ..density..), binwidth = 5, fill = "yellow", colour = "gray") +
  geom_density(colour = "red", size = 1) +
  ggtitle("Distribution of Infant Mortality Rate") +
  xlab("Infant Mortality Rate") +
  ylab("Density")
```

- Distribution of Infant Mortality Rate

The histogram shows that the distribution of infant_mortality_rate is right-skewed. This implies that most countries have low infant mortality rates, while a few have higher rates.


```{r, fig.height = 8}
# Income Class vs. Infant Mortality Rate
ggplot(data = sdg_clean, aes(x = income_class, y = infant_mortality_rate, fill = income_class)) +
  geom_boxplot(colour = "black") +
  scale_fill_manual(values = c("H" = "yellow", "UM" = "orange", "LM" = "red", "L" = "magenta")) +
  ggtitle("Income Class vs. Infant Mortality Rate") +
  xlab("Income Class") +
  ylab("Infant Mortality Rate")
```

```{r, fig.width = 12, fig.height = 12}
# Plot the infant mortality rate by country
ggplot(data = sdg_clean, aes(x = country, y = infant_mortality_rate, fill = country)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom") +
  labs(title = "Infant Mortality Rate by Country", x = "Country", y = "Infant Mortality Rate")
```

```{r}
# Show the top 5 countries with the lowest infant mortality rates and their respective income classes and vice versa
sdg_clean %>%
  arrange(infant_mortality_rate) %>%
  head(5) %>%
  dplyr::select(country, infant_mortality_rate, income_class) %>%
  kable(caption = "Top 5 countries with the lowest infant mortality rates") %>%
  kable_styling(full_width = F)
```

```{r}
# Show the bottom 5 countries with the highest infant mortality rates and their respective income classes
sdg_clean %>%
  arrange(desc(infant_mortality_rate)) %>%
  dplyr::select(country, infant_mortality_rate, income_class) %>%
  head(5) %>%
  kable(caption = "Top 5 countries with the highest infant mortality rates") %>%
  kable_styling(full_width = F)
```

- Infant Mortality Rate by Country

The bar plot displays the infant mortality rate for each country. There is a clear pattern in that as income class decrease from 'H' (upper) income class to 'L' (low) income class, the infant mortality rate tends to increase.

This is as shown that the table that the top 5 countries with the lowest infant mortality rates are all classified as 'H' (upper) income class countries. Conversely, the bottom 5 countries with the highest infant mortality rates are classified as 'L' (low) or 'LM" (low-middle)  income class countries.

```{r}
# Plot the relationship between infant mortality rate and years of education
ggplot(data = sdg_clean, aes(x = years_education, y = infant_mortality_rate)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = F, colour = "blue") +
  # geom_smooth(method = "loess", se = F, colour = "red") +
  ggtitle("Years of Education vs. Infant Mortality Rate") +
  xlab("Years of Education") +
  ylab("Infant Mortality Rate")
```

- Years of Education vs. Infant Mortality Rate

The scatter plot shows a negative relationship. Countries with higher average years of education tend to have lower infant mortality rates.

```{r, fig.width = 12, fig.height = 12}
# Life Expectancy vs. Infant Mortality Rate
ggplot(data = sdg_clean, aes(x = life_expectancy, y = infant_mortality_rate)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = F, colour = "blue") +
  # geom_smooth(method = "loess", se = F, colour = "red") +
  ggtitle("Life Expectancy vs. Infant Mortality Rate") +
  xlab("Life Expectancy") +
  ylab("Infant Mortality Rate")
```


- Life Expectancy vs. Infant Mortality Rate

The scatter plot shows a negative relationship. Countries with higher life expectancy tend to have lower infant mortality rates.

```{r, fig.width = 12, fig.height = 8}
# GDP vs. Infant Mortality Rate
ggplot(data = sdg_clean, aes(x = gdp, y = infant_mortality_rate)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, colour = "red", ) +
  ggtitle("GDP vs. Infant Mortality Rate") +
  xlab("GDP per Capita") +
  ylab("Infant Mortality Rate")

# GDP vs. Infant Mortality Rate for each continent
ggplot(data = sdg_clean, aes(x = gdp, y = infant_mortality_rate, colour = continent)) +
  geom_point(alpha = 0.5) +
  # geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~continent) +
  ggtitle("GDP vs. Infant Mortality Rate by Continent") +
  xlab("GDP per Capita") +
  ylab("Infant Mortality Rate")

# GDP vs. Infant Mortality Rate for each income class
ggplot(data = sdg_clean, aes(x = gdp, y = infant_mortality_rate, colour = income_class)) +
  geom_point(alpha = 0.5) +
  # geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~income_class) +
  ggtitle("GDP vs. Infant Mortality Rate by Income Class") +
  xlab("GDP per Capita") +
  ylab("Infant Mortality Rate")

```

- Life Expectancy vs. Infant Mortality Rate

The scatter plot shows a negative relationship. Countries with higher life expectancy tend to have lower infant mortality rates.

## Regression models

Linear Model:

$(E(\text{infant_mortality_rate}) = \beta_0 + \beta_1 \times \text{years_education} + \beta_2 \times \text{gdp} + \beta_3 \times \text{life_expectancy} + \beta_4 \times \text{income_class})$

```{r}
# Linear Model
lm_infant <- lm(infant_mortality_rate ~ years_education + gdp + life_expectancy + income_class, data = sdg_clean)
summary(lm_infant)
```










Log-Transformed Linear Model:

$E(\ln(\text{infant_mortality_rate})) = \beta_0 + \beta_1 \times \text{years_education} + \beta_2 \times \text{gdp} + \beta_3 \times \text{life_expectancy} + \beta_4 \times \text{income_class}$

```{r}
# Log-Transformed Linear Model
log_lm_infant <- lm(log(infant_mortality_rate) ~ years_education + gdp + life_expectancy + income_class, data = sdg_clean)
tidy(log_lm_infant)
```

```{r}
# Generate predictions for the log-transformed model
prediction_log_lm_infant <- predict(log_lm_infant)
prediction_log_lm_infant_exp <- exp(prediction_log_lm_infant)

ggplot(data = sdg_clean, aes(x = infant_mortality_rate, y = prediction_log_lm_infant_exp)) +
  geom_point(alpha = 0.5) +
  geom_abline(colour = "purple", size = 1, alpha = 0.4) +
  ggtitle("Log-Transformed Linear Model Predictions vs. Actual") +
  xlab("Actual Infant Mortality Rate") +
  ylab("Predicted Infant Mortality Rate")

# Effect Plot for Life Expectancy
ggplot(data = sdg_clean, aes(x = life_expectancy, y = infant_mortality_rate)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = prediction_log_lm_infant_exp), colour = "purple", alpha = 0.4) +
  ggtitle("Log-Transformed Linear Model Effect Plot - Life Expectancy") +
  xlab("Life Expectancy") +
  ylab("Infant Mortality Rate")

```


The plots for the log-transformed linear model indicate:

- Predictions vs. Actual: The first plot shows a strong alignment between the predicted and actual values, but there are some under-predictions for higher actual values.

- Effect Plot - Life Expectancy: The second plot shows a strong negative relationship between life_expectancy and infant_mortality_rate, similar to the GLM model.

Coefficients for the log-transformed linear model indicate that:
- Years of Education: For each additional year of education, the infant mortality rate decreases by 14.2% (exp⁡(−0.153)−1exp(−0.153)−1).
- Life Expectancy: For each additional year of life expectancy, the infant mortality rate decreases by 10.4% (exp⁡(−0.109)−1exp(−0.109)−1).
- Income Class: Lower income classes are associated with higher infant mortality rates, similar to the GLM model.


Generalised Linear Model (GLM):

$E(\text{infant_mortality_rate}) = \exp\left(\beta_0 + \beta_1 \times \text{years_education} + \beta_2 \times \text{gdp} + \beta_3 \times \text{life_expectancy} + \beta_4 \times \text{income_class}\right)$

```{r}
# GLM
glm_infant <- glm(infant_mortality_rate ~ years_education + gdp + life_expectancy + income_class, data = sdg_clean, family = poisson(link = "log"))
summary(glm_infant)
tidy(glm_infant)
```

```{r}
# Generate predictions for the GLM model
prediction_glm_infant <- predict(glm_infant, type = "response")

ggplot(data = sdg_clean, aes(x = infant_mortality_rate, y = prediction_glm_infant)) +
  geom_point(alpha = 0.5) +
  geom_abline(colour = "purple", size = 1, alpha = 0.4) +
  ggtitle("GLM Predictions vs. Actual") +
  xlab("Actual Infant Mortality Rate") +
  ylab("Predicted Infant Mortality Rate")

ggplot(data = sdg_clean, aes(x = life_expectancy, y = infant_mortality_rate)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = prediction_glm_infant), colour = "purple", alpha = 0.4) +
  ggtitle("GLM Effect Plot - Life Expectancy") +
  xlab("Life Expectancy") +
  ylab("Infant Mortality Rate")
```
The plots for the GLM model indicate:

- Predictions vs. Actual: The first plot shows that the predictions closely align with the actual infant mortality rates, but there are some deviations for higher values.

- Effect Plot - Life Expectancy: The second plot shows a strong negative relationship between life_expectancy and infant_mortality_rate, which aligns with our earlier observations.

Coefficients for the GLM model indicate that:

- Years of Education: For each additional year of education, the infant mortality rate decreases by 7.1% (exp⁡(−0.071)−1exp(−0.071)−1).
- Life Expectancy: For each additional year of life expectancy, the infant mortality rate decreases by 8.4% (exp⁡(−0.084)−1exp(−0.084)−1).
- Income Class: Lower income classes are associated with higher infant mortality rates, with the coefficients representing the log of the multiplicative effect compared to the "H" (high-income) category.


## Conclusion

Both the GLM and log-transformed linear models are effective in characterizing the behavior of infant mortality rate. The GLM model shows a slightly better fit based on the AIC value, but both models highlight the significant negative relationships with years of education and life expectancy, and the influence of income class. The findings suggest that investing in education and healthcare could have positive effects on reducing infant mortality rates across different countries and income classifications. ​







Your report here.

# Wine data

```{r}
# Use janitor to clean the data
wine_clean <- Wine %>%
  clean_names() %>%
  remove_empty(c("rows", "cols"))

# check unique values in the year column
# unique(wine_clean$year)
```
```{r}
# remove the year column with "N.V." values
wine_clean_year <- wine_clean %>%
  filter(year != "N.V.")
```

```{r}
# Distribution of wine ratings
ggplot(data = wine_clean_year, aes(x=rating)) +
  geom_histogram(binwidth=0.1, fill="yellow", color="gray", alpha=0.7) +
  labs(title="Distribution of Wine Ratings", x="Rating", y="Frequency")
```

Distribution of Wine Ratings shows a normal distribution with a peak around 3.8. This indicates that most wines have ratings around 3.8, with fewer wines having higher or lower ratings.

```{r}
# Plot the relationship between rating and wine variety
ggplot(wine_clean_year, aes(x = variety, y = rating)) +
  geom_boxplot(color = "black", aes(fill = variety)) +
  scale_fill_manual(values = c("White" = "yellow", "Sparkling" = "orange", 
                               "Red" = "red", "Rose" = "magenta")) +
  labs(title = "Ratings by Wine Variety", x = "Variety", y = "Rating")
```

```{r, fig.width = 12, fig.height = 8}
# Rating by country
ggplot(wine_clean_year, aes(x = country, y = rating)) +
  geom_boxplot(color = "black", aes(fill = country)) +
  labs(title = "Ratings by Country", x = "Country", y = "Rating") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom")
```

```{r}
# Explore how rating correlate with price of wine
ggplot(data = wine_clean_year, aes(x=price, y=rating)) +
  geom_point(alpha=0.3) +
  scale_x_log10() +
  labs(title="Ratings vs. Price", x="Price (Euros)", y="Rating")
```

```{r, fig.width = 12, fig.height = 5}
# Explore the ratings over the years
ggplot(data = wine_clean_year, aes(x=year, y=rating, colour=variety)) +
  geom_point(alpha=0.8) +
  geom_smooth(method="loess", se=FALSE) +
  labs(title="Ratings over the Years", x="Year", y="Rating") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
```


## Regression models

The first model will be a simple linear regression of 'Rating' on 'Price'. We use a logarithmic transformation on the price because of its skewed distribution.

$E(\text{Rating}) = \beta_0 + \beta_1 \times \log(\text{Price})$

```{r}
# Linear regression model
lm_wine <- lm(rating ~ log(price), data = wine_clean_year)
summary(lm_wine)
```

The linear regression model on 'Rating' using 'log(Price)' as a predictor can be expressed as

$E(\text{Rating}) = 3.131  + 0.249 \times \log(\text{Price})$

```{r}
# Generate predictions for the linear model
prediction_lm_wine <- predict(lm_wine)

ggplot(data = wine_clean_year, aes(x = rating, y = prediction_lm_wine)) +
  geom_point(alpha = 0.5) +
  # scale_x_log10() +
  # geom_abline(colour = "purple", size = 2, alpha = 0.4) +
  # geom_smooth(method = "lm", se = FALSE, colour = "blue") +
  geom_smooth(method = "loess", se = FALSE, colour = "red") +
  ggtitle("Linear Model Predictions vs. Actual") +
  xlab("Actual Rating") +
  ylab("Predicted Rating")
```

The plot shows the relationship between the actual and predicted ratings using the linear model. The red line represents the predicted values, which closely align with the actual data points. The LOESS model provides a good fit for the data, capturing the general trend of the relationship between price and rating.

Next, we consider a linear regression on 'Price' and 'NumberofRatings' as predictors of 'Rating'.

$E(\text{Rating}) = \beta_0 + \beta_1 \times \log(\text{Price}) + \beta_2 \times \log(\text{NumberofRatings})$

```{r}
# Linear regression model with Price and NumberofRatings as predictors
lm_wine_price_ratings <- lm(rating ~ log(price) + log(number_of_ratings), data = wine_clean_year)

summary(lm_wine_price_ratings)
```

The linear regression model on 'Rating' using 'log(Price)' and 'log(NumberofRatings)' as predictors can be expressed as

$E(\text{Rating}) = 3.044 + 0.248 \times \log(\text{Price}) + 0.018 \times \log(\text{NumberofRatings})$

```{r}
# Generate predictions for the linear model with Price and NumberofRatings as predictors
prediction_lm_wine_price_ratings <- predict(lm_wine_price_ratings)

ggplot(data = wine_clean_year, aes(x = rating, y = prediction_lm_wine_price_ratings)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, colour = "red") +
  ggtitle("Linear Model with Price and Number of Ratings Predictions vs. Actual") +
  xlab("Actual Rating") +
  ylab("Predicted Rating")
```

```{r}
# Obtain the AIC values for the models
AIC(lm_wine, lm_wine_price_ratings)
```
Given the similarity in AIC values for the two models, the model with 'Price' and 'NumberofRatings' as predictors does not provide a significant improvement over the simpler model with only 'Price' as a predictor. The simpler model may be preferred due to its parsimony and similar predictive performance.

Therefore, the linear regression model with 'log(Price)' as a predictor provides a good fit for the data and captures the relationship between price and rating effectively.















































# References

```{=html}
<!--
Include your references in this section. The ones below are just examples.
-->
```
1.  Wickham, H. *ggplot2: Elegant Graphics for Data Analysis* [on-line version](https://ggplot2-book.org/index.html)

2.  Wickham, H. & Grolemund, G. *R for Data Science.* [on-line version](https://r4ds.had.co.nz/index.html)

3.  [Department of Maths website](https://www.bath.ac.uk/departments/department-of-mathematical-sciences/)

```{r}
```

```{r}
# Check the structure of the data
str(sdg_clean)
```

```{r}
# Summarise the data
summary(sdg_clean)
```



```{r, fig.width = 12, fig.height = 12}
ggplot(data = sdg_clean, aes(x = years_education, y = infant_mortality_rate, colour = country)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom") +
  labs(title = "Infant Mortality Rate by Years of Education", x = "Years of Education", y = "Infant Mortality Rate")
```

```{r, fig.width = 12, fig.height = 12}
glm_education_year_vs_infant_mortality_rate <- ggplot(
      data = sdg_clean,
      aes(x = years_education,
          y = infant_mortality_rate,
          colour = country)) +
      geom_point() +
      geom_smooth(
        method = "glm", 
        method.args = list(family = binomial(link="logit")), 
        formula = y ~ x, 
        se = FALSE, 
        linewidth = 0.5
      ) +
      theme(axis.text.x = element_text(angle = 90, hjust = 1), 
            legend.position = "bottom") +
      labs(title = "Infant Mortality Rate by Years of Education", 
           x = "Years of Education", 
           y = "Infant Mortality Rate")
glm_education_year_vs_infant_mortality_rate

# Infant Mortality Rate vs. Years of Education:
# 
#     There is a general downward trend suggesting that as the average years of education increases, the infant mortality rate tends to decrease.
#     The trend seems reasonably linear.
```

```{r, fig.width = 12, fig.height = 12}
ggplot(data = sdg_clean, aes(x = gdp, y = infant_mortality_rate, colour = country)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom") +
  labs(title = "Infant Mortality Rate by GDP", x = "GDP", y = "Infant Mortality Rate")

# Infant Mortality Rate vs. GDP per Capita:
# 
#     There is a significant downward trend indicating that higher GDP per capita correlates with lower infant mortality rates.
#     The trend is pronounced and suggests a potential exponential relationship.
```

```{r, fig.width = 12, fig.height = 12}
ggplot(data = sdg_clean, aes(x = life_expectancy, y = infant_mortality_rate, colour = country)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom") +
  labs(title = "Infant Mortality Rate by Life Expectancy", x = "Life Expectancy", y = "Infant Mortality Rate")

# Infant Mortality Rate vs. Life Expectancy:
# 
#     There is a clear downward trend indicating that countries with higher life expectancy generally have lower infant mortality rates.
#     The trend looks linear, though with some variability.
```

```{r, fig.width = 12, fig.height = 12}
ggplot(data = sdg_clean, aes(x = year, y = income_class, colour = country)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom") +
  labs(title = "Income Class by Year", x = "Year", y = "Income Class")
```

```{r, fig.width = 12, fig.height = 12}
ggplot(data = sdg_clean, aes(x = year, y = income_class, colour = country)) +
  geom_point() +
  # geom_smooth() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom") +
  labs(title = "Income Class by Year", x = "Year", y = "Income Class")
```

```{r}
# Model 1: Linear Regression for Infant Mortality Rate vs. Years of Education
model_education_year_vs_infant_mortality_rate <- lm(infant_mortality_rate ~ years_education, data = sdg_clean)
summary(model_education_year_vs_infant_mortality_rate)
```

```{r}
# Model 2: Log-Transformed Linear Regression (Infant Mortality Rate vs. GDP per Capita)
model_gdp_vs_infant_mortality_rate <- lm(log(infant_mortality_rate) ~ log(gdp), data = sdg_clean)
summary(model_gdp_vs_infant_mortality_rate)
```

```{r}
# Model 3: Linear Regression for Infant Mortality Rate vs. Life Expectancy
model_life_expectancy_vs_infant_mortality_rate <- lm(infant_mortality_rate ~ life_expectancy, data = sdg_clean)
summary(model_life_expectancy_vs_infant_mortality_rate)
```

```{r}
# Create plots for the models
ggplot(data = sdg_clean, aes(x = years_education, y = infant_mortality_rate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Infant Mortality Rate vs. Years of Education", x = "Years of Education", y = "Infant Mortality Rate")
```

```{r}
ggplot(data = sdg_clean, aes(x = log(gdp), y = log(infant_mortality_rate))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Infant Mortality Rate vs. GDP per Capita", x = "Log(GDP)", y = "Log(Infant Mortality Rate)")

# The first effect plot showcases the relationship between infant_mortality_rate and gdp using Model 2. The plot is on a log scale, and it reveals a strong negative correlation between GDP per capita and infant mortality rate. The red line illustrates the model's predicted values, closely fitting the actual data points.
# Model 2: Log-Transformed Linear Regression for infant_mortality_rate vs. gdp
# 
#     The conditional expectation: E[log⁡(Y)∣X]=4.3785−0.5962⋅log⁡(gdp)E[log(Y)∣X]=4.3785−0.5962⋅log(gdp)
#     This model provides a strong fit with an R2R2 of 0.777 and the lowest AIC value of 394.6.
#     This suggests that an exponential decay in infant mortality rate is associated with increasing GDP.
```

```{r}
ggplot(data = sdg_clean, aes(x = life_expectancy, y = infant_mortality_rate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Infant Mortality Rate vs. Life Expectancy", x = "Life Expectancy", y = "Infant Mortality Rate")

#The second effect plot demonstrates the relationship between infant_mortality_rate and life_expectancy using Model 3. The negative correlation is evident, with the red line depicting the predicted values. The model aligns well with the data points, showing that higher life expectancy generally corresponds to lower infant mortality rates.
# Model 3: Linear Regression for infant_mortality_rate vs. life_expectancy
# 
#     The conditional expectation: E[Y∣X]=176.7136−2.1573⋅life_expectancyE[Y∣X]=176.7136−2.1573⋅life_expectancy
#     This model provides a strong fit with an R2R2 of 0.783 and a low AIC value of 1708.0.
#     This suggests that higher life expectancy generally corresponds to lower infant mortality rates.

```

```{r}
# GPT conclusions for model 2 and 3
# Both models indicate strong negative relationships between infant mortality rates and the respective predictors (gdp and life_expectancy). The models highlight the influence of socio-economic and health-related factors on infant mortality. The GDP-based model showcases the impact of economic conditions, while the life expectancy model highlights the impact of health and quality of life on infant mortality.
```


```{r}
# write_csv(wine_clean_year, here("data", "wine_clean_year.csv"))
# write_csv(sdg_clean, here("data", "sdg_clean.csv"))
# write_csv(wine_clean, here("data", "wine_clean.csv"))
```

\`\`\`{r}
